Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-08-08 submission/manuscript.Rmd
female edits

@@

It is currently unknown if the misexpression of germline genes is influenced by chromosomal sex, as previous studies on germline gene repression have been conducted exclusively in males. Sex is particularly pertinent in the case of KDM5C, as KDM5C is more highly expressed in females than males. We assessed germline gene misexpression in _Kdm5c_ mutant male and female mouse epiblast-like cells (EpiLCs), an _in vitro_ model of the post-implantation embryo that reflects when germline genes are typically silenced in somatic cells.

We first identified differentially expressed genes (DEGs) compared to wild-type controls (DESeq2[@loveModeratedEstimationFold2014a], log2 fold change > `r l2fcco_ESCEpiLC`, q < 0.1) in male _Kdm5c_ hemizygous knockout (XY _Kdm5c_-KO), female homozygous knockout (XX _Kdm5c_-KO) and female heterozygous (XX _Kdm5c_-HET) EpiLCs and then filtered for germline-enriched genes using our previously curated list. While both male and female _Kdm5c_ mutants misexpressed germline genes, about twice as many were dysregulated in homozygous and heterozygous females compared to homozygous males (XX KO: , XX HET: , XY KO: ) (Figure 1A).

>>>>

- Add females - KDM5C is a sexually dimorphic chromatin regulator, embryonic lethality makes it difficult to compare

@@

It is currently unknown if the misexpression of germline genes is influenced by chromosomal sex, as previous studies on germline gene repressors have been conducted exclusively in males. We explored the impact of sex upon germline gene misexpession by comparing their dysregulation in male _Kdm5c_ hemizygous knockout (XY _Kdm5c_-KO), female homozygous knockout (XX _Kdm5c_-KO) and female heterozygous (XX _Kdm5c_-HET) EpiLCs. [@samantaActivationXistEvolutionarily2022].

We first identified differentially expressed genes (DEGs) compared to sex-matched wild-type controls (DESeq2[@loveModeratedEstimationFold2014a], log2 fold change > `r l2fcco_ESCEpiLC`, q < 0.1) and then filtered for germline-enriched genes.Homozygous and heterozygous females expressed over double the number of germline-enriched genes than hemizygous males (Figure 4A). While the majority of germline DEGs in _Kdm5c_-KO males were also dysregulated in females (74%), there were also many male-specific and female-specific germline DEGs (Figure 4A-B). We compared the known functions of germline genes dysregulated in all samples (shared), only in females (XX only - XX _Kdm5c_-KO, XX _Kdm5c_-HET, or both), and only in males (XY only).

- Gene ontologies uniquely enriched in female-specific germline DEGs included meiotic cell cycle and cilium movement, while mitochondrial and cell signaling gene ontologies were enriched in male-specifc DEGs (protein localization to mitochondrion - )
- XY only - protein localization to mitochondrion
- Shared germline DEGs
	- meiotic and 
- Shared germline genes included female gamete generation
